---
title: "Beaches_Bikes_Dwell"
author: "Derek Kusmenko"
format: pdf
editor: visual
---

## Importing Data and Libraries

```{r}
library(opendatatoronto)
library(tidyverse)
library(sf)
library(leaflet)
library(skimr)
library(janitor)
library(lubridate)
library(tidytext)
set.seed(2453)
```

```{r}
df <- read.csv("../01_Data/beaches_bikes.csv")
```

## Dwell Time Analysis (Weekends)

```{r}
# prepare data
flow_data <- df %>%
  # filter for weekends only
  filter(wday(start_date_actual, week_start = 1) %in% c(6, 7)) %>%
  
  # determine trip direction
  mutate(
    start_in_beaches = Start.Station.Id %in% beaches_stations,
    end_in_beaches   = End.Station.Id %in% beaches_stations,
    
    direction = case_when(
      !start_in_beaches & end_in_beaches  ~ "Inbound",  
      start_in_beaches & !end_in_beaches  ~ "Outbound",  
      TRUE ~ "Other"
    )
  ) %>%
  
  # filter out "Other" trips
  filter(direction %in% c("Inbound", "Outbound")) %>%
  
  # extract hour and trip count
  mutate(hour = hour(as.POSIXct(Start.Time, format = "%m/%d/%Y %H:%M"))) %>%
  group_by(direction, hour) %>%
  summarise(total_trips = n(), .groups = "drop")

# plot flow
ggplot(flow_data, aes(x = hour, y = total_trips, color = direction, group = direction)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +

  geom_area(aes(fill = direction), alpha = 0.2, position = "identity") +
  
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  
  labs(
    title = "Weekend Flow: Inbound vs. Outbound",
    subtitle = "Hourly Trip Volume (Saturdays & Sundays)",
    x = "Hour of Day",
    y = "Total Trips",
    color = "Flow Direction",
    fill = "Flow Direction"
  ) +
  theme_minimal()

```

```{r}
# reuse the filtering logic from above
dwell_analysis <- df %>%
  filter(wday(start_date_actual, week_start = 1) %in% c(6, 7)) %>%

  mutate(
    start_in_beaches = Start.Station.Id %in% beaches_stations,
    end_in_beaches   = End.Station.Id %in% beaches_stations,
    
    direction = case_when(
      !start_in_beaches & end_in_beaches  ~ "Inbound",
      start_in_beaches & !end_in_beaches  ~ "Outbound",
      TRUE ~ "Other"
    )
  ) %>%
  filter(direction %in% c("Inbound", "Outbound")) %>%
  mutate(hour = hour(as.POSIXct(Start.Time, format = "%m/%d/%Y %H:%M")))

# calculate weighted average starting hour 
dwell_summary <- dwell_analysis %>%
  group_by(direction) %>%
  summarise(
    # weighted avg: Sum(Hour * Count) / Total Count
    avg_hour_of_day = mean(hour),
    
    median_hour_of_day = median(hour),
    
    total_trips = n()
  )

print(dwell_summary)

# calculate difference in means
inbound_avg <- dwell_summary$avg_hour_of_day[dwell_summary$direction == "Inbound"]
outbound_avg <- dwell_summary$avg_hour_of_day[dwell_summary$direction == "Outbound"]

estimated_dwell_time <- outbound_avg - inbound_avg

print(paste("Estimated Average Dwell Time:", round(estimated_dwell_time, 2), "hours"))



```

```{r}
dwell_stats <- dwell_analysis %>%
  group_by(direction) %>%
  summarise(
    # use median instead of mean
    median_hour = median(hour),
    
    # find peak hour
    peak_hour = as.numeric(names(which.max(table(hour)))),
    
    # count total trip volume
    total_volume = n()
  )

# extract values
inbound_median <- dwell_stats$median_hour[dwell_stats$direction == "Inbound"]
outbound_median <- dwell_stats$median_hour[dwell_stats$direction == "Outbound"]

inbound_vol <- dwell_stats$total_volume[dwell_stats$direction == "Inbound"]
outbound_vol <- dwell_stats$total_volume[dwell_stats$direction == "Outbound"]

# calculate Metrics
dwell_time_median <- outbound_median - inbound_median
dropoff_rate <- (inbound_vol - outbound_vol) / inbound_vol

# print other metrics
print(paste("Typical Arrival Time (Median):", inbound_median, ":00"))
print(paste("Typical Departure Time (Median):", outbound_median, ":00"))
print(paste("Estimated Dwell Time:", dwell_time_median, "hours"))
print(paste("One-Way Trip Percentage:", round(dropoff_rate * 100, 1), "% of riders do not bike back."))


```

### Only Using Leisure Cluster data

Now repeating the analysis but using the data calculated from the leisure dataset generated in the _cluster file

```{r}

df2 <- read_csv("../01_Data/leisure_dataset.csv")

# redo analysis from above
flow_data2 <- df2 %>%
  filter(wday(start_date_actual, week_start = 1) %in% c(6, 7)) %>%
  mutate(
    start_in_beaches = Start.Station.Id %in% beaches_stations,
    end_in_beaches   = End.Station.Id %in% beaches_stations,
    direction = case_when(
      !start_in_beaches & end_in_beaches  ~ "Inbound",   
      start_in_beaches & !end_in_beaches  ~ "Outbound",  
      TRUE ~ "Other" 
    )
  ) %>%
  filter(direction %in% c("Inbound", "Outbound")) %>%
  mutate(hour = hour(as.POSIXct(Start.Time, format = "%m/%d/%Y %H:%M"))) %>%
  group_by(direction, hour) %>%
  summarise(total_trips = n(), .groups = "drop")

# plot the flow
ggplot(flow_data2, aes(x = hour, y = total_trips, color = direction, group = direction)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  
  geom_area(aes(fill = direction), alpha = 0.2, position = "identity") +
  
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  
  labs(
    title = "Weekend Flow: Inbound vs. Outbound",
    subtitle = "Hourly Trip Volume (Saturdays & Sundays)",
    x = "Hour of Day",
    y = "Total Trips",
    color = "Flow Direction",
    fill = "Flow Direction"
  ) +
  theme_minimal()

```

```{r}
# again reusing logic

dwell_analysis2 <- df2 %>%
  filter(wday(start_date_actual, week_start = 1) %in% c(6, 7)) %>%
  mutate(
    start_in_beaches = Start.Station.Id %in% beaches_stations,
    end_in_beaches   = End.Station.Id %in% beaches_stations,
    
    direction = case_when(
      !start_in_beaches & end_in_beaches  ~ "Inbound",
      start_in_beaches & !end_in_beaches  ~ "Outbound",
      TRUE ~ "Other"
    )
  ) %>%
  filter(direction %in% c("Inbound", "Outbound")) %>%

  mutate(hour = hour(as.POSIXct(Start.Time, format = "%m/%d/%Y %H:%M")))

dwell_summary2 <- dwell_analysis2 %>%
  group_by(direction) %>%
  summarise(
    avg_hour_of_day = mean(hour),
    median_hour_of_day = median(hour),
    total_trips = n()
  )

print(dwell_summary2)

# calculate differences
inbound_avg2 <- dwell_summary2$avg_hour_of_day[dwell_summary2$direction == "Inbound"]
outbound_avg2 <- dwell_summary2$avg_hour_of_day[dwell_summary2$direction == "Outbound"]

estimated_dwell_time2 <- outbound_avg2 - inbound_avg2

print(paste("Estimated Average Dwell Time:", round(estimated_dwell_time2, 2), "hours"))



```

```{r}
# again using same code
dwell_stats2 <- dwell_analysis2 %>%
  group_by(direction) %>%
  summarise(
    median_hour = median(hour),
    peak_hour = as.numeric(names(which.max(table(hour)))),
    total_volume = n()
  )

inbound_median2 <- dwell_stats2$median_hour[dwell_stats2$direction == "Inbound"]
outbound_median2 <- dwell_stats2$median_hour[dwell_stats2$direction == "Outbound"]

inbound_vol2 <- dwell_stats2$total_volume[dwell_stats2$direction == "Inbound"]
outbound_vol2 <- dwell_stats2$total_volume[dwell_stats2$direction == "Outbound"]

# calculate Metrics
dwell_time_median2 <- outbound_median2 - inbound_median2
dropoff_rate2 <- (inbound_vol2 - outbound_vol2) / inbound_vol2

# print metrics
print(paste("Typical Arrival Time (Median):", inbound_median2, ":00"))
print(paste("Typical Departure Time (Median):", outbound_median2, ":00"))
print(paste("Estimated Dwell Time:", dwell_time_median2, "hours"))
print(paste("One-Way Trip Percentage:", round(dropoff_rate2 * 100, 1), "% of riders do not bike back."))


```

## Repeat for all days of the week (not just weekends)

```{r}
# use same code as above but removing weekend filter
flow_data3 <- df %>%

  mutate(
    start_in_beaches = Start.Station.Id %in% beaches_stations,
    end_in_beaches   = End.Station.Id %in% beaches_stations,
    
    direction = case_when(
      !start_in_beaches & end_in_beaches  ~ "Inbound",   
      start_in_beaches & !end_in_beaches  ~ "Outbound",  
      TRUE ~ "Other" 
    )
  ) %>%

  filter(direction %in% c("Inbound", "Outbound")) %>%
  
  mutate(hour = hour(as.POSIXct(Start.Time, format = "%m/%d/%Y %H:%M"))) %>%
  group_by(direction, hour) %>%
  summarise(total_trips = n(), .groups = "drop")

# plot
ggplot(flow_data3, aes(x = hour, y = total_trips, color = direction, group = direction)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +

  geom_area(aes(fill = direction), alpha = 0.2, position = "identity") +
  
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  
  labs(
    title = "Bike Flow: Inbound vs. Outbound",
    subtitle = "Hourly Trip Volume",
    x = "Hour of Day",
    y = "Total Trips",
    color = "Flow Direction",
    fill = "Flow Direction"
  ) +
  theme_minimal()

```

```{r}
# reusing code
dwell_analysis3 <- df %>%

  mutate(
    start_in_beaches = Start.Station.Id %in% beaches_stations,
    end_in_beaches   = End.Station.Id %in% beaches_stations,
    
    direction = case_when(
      !start_in_beaches & end_in_beaches  ~ "Inbound",
      start_in_beaches & !end_in_beaches  ~ "Outbound",
      TRUE ~ "Other"
    )
  ) %>%
  filter(direction %in% c("Inbound", "Outbound")) %>%

  mutate(hour = hour(as.POSIXct(Start.Time, format = "%m/%d/%Y %H:%M")))

dwell_summary3 <- dwell_analysis3 %>%
  group_by(direction) %>%
  summarise(

    avg_hour_of_day = mean(hour),

    median_hour_of_day = median(hour),
    
    total_trips = n()
  )

print(dwell_summary3)

inbound_avg3 <- dwell_summary3$avg_hour_of_day[dwell_summary3$direction == "Inbound"]
outbound_avg3 <- dwell_summary3$avg_hour_of_day[dwell_summary3$direction == "Outbound"]

estimated_dwell_time3 <- outbound_avg3 - inbound_avg3

print(paste("Estimated Average Dwell Time:", round(estimated_dwell_time3, 2), "hours"))
```

```{r}
# reusing code
dwell_stats3 <- dwell_analysis3 %>%
  group_by(direction) %>%
  summarise(

    median_hour = median(hour),

    peak_hour = as.numeric(names(which.max(table(hour)))),

    total_volume = n()
  )

inbound_median3 <- dwell_stats3$median_hour[dwell_stats3$direction == "Inbound"]
outbound_median3 <- dwell_stats3$median_hour[dwell_stats3$direction == "Outbound"]

inbound_vol3 <- dwell_stats3$total_volume[dwell_stats3$direction == "Inbound"]
outbound_vol3<- dwell_stats3$total_volume[dwell_stats3$direction == "Outbound"]


dwell_time_median3 <- outbound_median3 - inbound_median3
dropoff_rate3 <- (inbound_vol3 - outbound_vol3) / inbound_vol3

print(paste("Typical Arrival Time (Median):", inbound_median3, ":00"))
print(paste("Typical Departure Time (Median):", outbound_median3, ":00"))
print(paste("Estimated Dwell Time:", dwell_time_median3, "hours"))
print(paste("One-Way Trip Percentage:", round(dropoff_rate3 * 100, 1), "% of riders do not bike back."))


```

### Only Using Leisure Cluster Data

```{r}
# reusing code
flow_data4 <- df2 %>%

  mutate(
    start_in_beaches = Start.Station.Id %in% beaches_stations,
    end_in_beaches   = End.Station.Id %in% beaches_stations,
    
    direction = case_when(
      !start_in_beaches & end_in_beaches  ~ "Inbound",   
      start_in_beaches & !end_in_beaches  ~ "Outbound",  
      TRUE ~ "Other" 
    )
  ) %>%

  filter(direction %in% c("Inbound", "Outbound")) %>%

  mutate(hour = hour(as.POSIXct(Start.Time, format = "%m/%d/%Y %H:%M"))) %>%
  group_by(direction, hour) %>%
  summarise(total_trips = n(), .groups = "drop")

ggplot(flow_data4, aes(x = hour, y = total_trips, color = direction, group = direction)) +
  geom_line(size = 1.5) +
  geom_point(size = 3) +
  geom_area(aes(fill = direction), alpha = 0.2, position = "identity") +
  
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  
  labs(
    title = "Bike Flow: Inbound vs. Outbound",
    subtitle = "Hourly Trip Volume ",
    x = "Hour of Day",
    y = "Total Trips",
    color = "Flow Direction",
    fill = "Flow Direction"
  ) +
  theme_minimal()

```

```{r}
# reuse code
dwell_analysis4 <- df2 %>%
  
  mutate(
    start_in_beaches = Start.Station.Id %in% beaches_stations,
    end_in_beaches   = End.Station.Id %in% beaches_stations,
    
    direction = case_when(
      !start_in_beaches & end_in_beaches  ~ "Inbound",
      start_in_beaches & !end_in_beaches  ~ "Outbound",
      TRUE ~ "Other"
    )
  ) %>%
  filter(direction %in% c("Inbound", "Outbound")) %>%
  
  mutate(hour = hour(as.POSIXct(Start.Time, format = "%m/%d/%Y %H:%M")))


dwell_summary4 <- dwell_analysis4 %>%
  group_by(direction) %>%
  summarise(

    avg_hour_of_day = mean(hour),

    median_hour_of_day = median(hour),
    
    total_trips = n()
  )

print(dwell_summary4)


inbound_avg4 <- dwell_summary4$avg_hour_of_day[dwell_summary4$direction == "Inbound"]
outbound_avg4<- dwell_summary4$avg_hour_of_day[dwell_summary4$direction == "Outbound"]

estimated_dwell_time4 <- outbound_avg4 - inbound_avg4

print(paste("Estimated Average Dwell Time:", round(estimated_dwell_time4, 2), "hours"))
```

```{r}
dwell_stats4 <- dwell_analysis4 %>%
  group_by(direction) %>%
  summarise(

    median_hour = median(hour),

    peak_hour = as.numeric(names(which.max(table(hour)))),

    total_volume = n()
  )

inbound_median4 <- dwell_stats4$median_hour[dwell_stats4$direction == "Inbound"]
outbound_median4 <- dwell_stats4$median_hour[dwell_stats4$direction == "Outbound"]

inbound_vol4 <- dwell_stats4$total_volume[dwell_stats4$direction == "Inbound"]
outbound_vol4 <- dwell_stats4$total_volume[dwell_stats4$direction == "Outbound"]

dwell_time_median4 <- outbound_median4 - inbound_median4
dropoff_rate4 <- (inbound_vol4 - outbound_vol4) / inbound_vol4

print(paste("Typical Arrival Time (Median):", inbound_median4, ":00"))
print(paste("Typical Departure Time (Median):", outbound_median4, ":00"))
print(paste("Estimated Dwell Time:", dwell_time_median4, "hours"))
print(paste("One-Way Trip Percentage:", round(dropoff_rate4 * 100, 1), "% of riders do not bike back."))


```

See after including all days, filtering for leisure cluster only makes data look more like weekend only, again indicating leisure cluster truly includes mainly leisure trips
