---
title: "Beaches_Bikes_Cluster"
author: "Derek Kusmenko"
format: pdf
editor: visual
---

## Importing Data and Libraries

```{r}
library(opendatatoronto)
library(tidyverse)
library(sf)
library(leaflet)
library(skimr)
library(janitor)
library(lubridate)
library(tidytext)
library(readr)
library(dplyr)
library(ggplot2)
library(purrr)
library(leaflet)
library(dbscan)
library(igraph)
library(tidygraph)
library(ggraph)
```

```{r}
trips_data <- read_csv("../01_Data/beaches_bikes.csv")
```

## K-means

First want to plot a scree plot to use the "Elbow Method" to find the optimal number of clusters.

```{r}
set.seed(2453)
kmeans_data <- trips_data |>
  select(Trip..Duration, start_hour_of_day, User.Type) |>
  filter(!is.na(Trip..Duration) & !is.na(start_hour_of_day) & !is.na(User.Type)) |>
  filter(Trip..Duration < 7200) # Remove massive outliers so clusters aren't skewed

# scale the data
kmeans_scaled <- scale(kmeans_data |> select(Trip..Duration, start_hour_of_day))

# calculate for k = 1 to 10
k_values <- 1:10

wss_values <- map_dbl(k_values, function(k) {
  kmeans(kmeans_scaled, centers = k, nstart = 10)$tot.withinss
})

# create dataframe to store values for plotting
scree_data <- data.frame(
  k = k_values,
  WSS = wss_values
)

# plot the Scree Plot
ggplot(scree_data, aes(x = k, y = WSS)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkred", size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Scree Plot (Elbow Method) for Trip Profiles",
    subtitle = "Look for the 'bend' or 'elbow' to find the optimal k",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster Sum of Squares"
  ) +
  theme_minimal()
```

So 3 clusters appears to be the optimal number via the Elbow Method. Proceed with k=3 means clustering.

```{r}
set.seed(2453)
# Run K-Means with 3 clusters
trip_clusters <- kmeans(kmeans_scaled, centers = 3, nstart = 25)

# Add the cluster assignments back to the data
kmeans_data$Cluster <- as.factor(trip_clusters$cluster)

# Visualize the Trip Profiles
ggplot(kmeans_data, aes(x = start_hour_of_day, y = Trip..Duration, color = Cluster)) +
  geom_point(alpha = 0.3) +
  labs(title = "Trip Profiles (K-Means)",
       x = "Hour of Day",
       y = "Trip Duration (Seconds)") +
  theme_minimal()
```

From this plot we see that are 3 clusters are broken down as early trips, late trips, and long trips. We can further analyze the data in each cluster.

```{r}
# Summarise the profiles including User Type Proportions
cluster_profiles <- kmeans_data |>
  group_by(Cluster) |>
  summarise(
    Total_Trips = n(),
    Avg_Start_Hour = round(mean(start_hour_of_day), 1),
    Avg_Duration_Mins = round(mean(Trip..Duration) / 60, 1),
    
    # Calculate the percentage of each member type
    Pct_Annual = round(sum(User.Type == "Annual Member") / n() * 100, 1),
    Pct_Casual = round(sum(User.Type == "Casual Member") / n() * 100, 1)
  ) |>
  arrange(Cluster)

# Print the resulting table
print(cluster_profiles)
```

From this output, it appears that cluster 1 mainly comprises of longer trips all throughout the day, with the average start time at about 3:30pm. We also have a very large concentration of casual to annual members, indicating that this cluster is likely a good representation of leisure trips. This is a strong assumption as we are likely missing other leisure trips that are being captured by the other clusters, however, it is a sizeable sample to proceed with.

```{r}
set.seed(2453)
# list of id's for stations in Beaches BIA
beaches_stations <- c(
  "7692",
  "7317",
  "7318",
  "7365",
  "7427",
  "7315",
  "7695",
  "7364",
  "7314",
  "7316",
  "7428"
)

# Prepare and Filter the Data for INBOUND trips
kmeans_data_inbound <- trips_data |>
  # Filter for trips ENDING in Beaches
  filter(End.Station.Id %in% beaches_stations) |>
  filter(!(Start.Station.Id %in% beaches_stations)) |>
  # Select Duration, END Hour, and User Type
  select(Trip..Duration, end_hour_of_day, User.Type) |>
  filter(!is.na(Trip..Duration) & !is.na(end_hour_of_day) & !is.na(User.Type)) |>
  filter(Trip..Duration < 7200) # Remove massive outliers

# Scale the numerical columns (Duration and End Hour)
kmeans_scaled_inbound <- scale(kmeans_data_inbound |> select(Trip..Duration, end_hour_of_day))

# Run K-Means with 3 clusters
inbound_clusters <- kmeans(kmeans_scaled_inbound, centers = 3, nstart = 25)

# Add the cluster assignments back to the original unscaled data
kmeans_data_inbound$Cluster <- as.factor(inbound_clusters$cluster)

# Summarise the Inbound Profiles
inbound_profiles <- kmeans_data_inbound |>
  group_by(Cluster) |>
  summarise(
    Total_Trips = n(),
    Avg_Arrival_Hour = round(mean(end_hour_of_day), 1),
    Avg_Duration_Mins = round(mean(Trip..Duration) / 60, 1),
    
    # Calculate member type proportions
    Pct_Annual = round(sum(User.Type == "Annual Member") / n() * 100, 1),
    Pct_Casual = round(sum(User.Type == "Casual Member") / n() * 100, 1)
  ) |>
  arrange(Cluster)

# Print the resulting table
print(inbound_profiles)
```

See similar results for inbound only trips as all trips (now cluster 2)

### Create Cluster 1 Only Dataset

```{r}
set.seed(2453)
# load data
trips_data <- read_csv("../01_Data/beaches_bikes.csv")

# remove any na and extreme outliers
df_cleaned <- trips_data |>
  filter(!is.na(Trip..Duration) & !is.na(start_hour_of_day)) |>
  filter(Trip..Duration < 7200) # Keep trips under 2 hours to avoid skew

# scale and Run K-Means (k = 3)
kmeans_scaled <- scale(df_cleaned |> select(Trip..Duration, start_hour_of_day))
trip_clusters <- kmeans(kmeans_scaled, centers = 3, nstart = 25)

# add Cluster Assignments back to the cleaned dataset
df_cleaned$Cluster <- trip_clusters$cluster


# to keep track of which cluster is for leisure as sometimes label changes when re running
leisure_cluster <- which(df_cleaned |> 
                          group_by(Cluster) |>
                          summarize(n = n()) == 27009) - 3

# filter for leisure cluster
leisure_dataset <- df_cleaned |>
  filter(Cluster == leisure_cluster)

# save to csv for future use
write_csv(leisure_dataset, "../01_Data/leisure_dataset.csv")

# summary
cat("Export Complete!\n")
cat("Total trips in Leisure Cluster:", nrow(leisure_dataset), "\n")
cat("Average Duration (Mins):", round(mean(leisure_dataset$Trip..Duration)/60, 1), "\n")
cat("Average Start Hour:", round(mean(leisure_dataset$start_hour_of_day), 1), "\n")
```

## DBScan

```{r}
set.seed(2453)
# load in station data for lat and long
stations_data <- read_csv("../01_Data/Bike_Station_Info.csv")
stations_data <- stations_data |> mutate(station_id = as.character(station_id))

# find top 50 most common trips
top_50_trips <- trips_data |>
  filter(!is.na(Start.Station.Id) & !is.na(End.Station.Id)) |>
  group_by(Start.Station.Id, End.Station.Id) |>
  summarise(weight = n(), .groups = "drop") |>
  filter(Start.Station.Id != End.Station.Id) |>
  arrange(desc(weight)) |>
  slice_head(n = 50)

#create list of top stations
top_stations_list <- unique(c(top_50_trips$Start.Station.Id, top_50_trips$End.Station.Id))

# filter to only include above stations
spatial_data_top50 <- stations_data |>
  filter(station_id %in% top_stations_list) |>
  select(station_id, name, lat, lon) |>
  filter(!is.na(lat) & !is.na(lon))

# Extract just the coordinates for the DBSCAN algorithm
coords_only <- spatial_data_top50 |> select(lon, lat)

# Run DBSCAN 
dbscan_res <- dbscan(coords_only, eps = 0.01, minPts = 2)

# Add the cluster assignments back to the spatial data
spatial_data_top50$Cluster <- as.factor(dbscan_res$cluster)

# plot results on map of Toronto
pal_clusters <- colorFactor(palette = "Set1", domain = spatial_data_top50$Cluster)

# Generate the Map
leaflet(spatial_data_top50) |>
  addTiles() |>
  setView(lng = -79.3832, lat = 43.6532, zoom = 12) |>
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    radius = 8, 
    fillColor = ~pal_clusters(Cluster),
    fillOpacity = 0.9,
    stroke = TRUE,
    weight = 1,
    color = "black", 
    popup = ~paste("<b>Station:</b>", name, "<br>",
                   "<b>Top 50 Spatial Cluster:</b>", Cluster)
  ) |>
  addLegend("bottomright", pal = pal_clusters, values = ~Cluster,
            title = "Top 50 Spatial Clusters<br>(0 = Outliers)",
            opacity = 1)
```

## Network Graphs

```{r}
set.seed(2453)
# Create edges for top 100 trips
edges <- trips_data |>
  filter(!is.na(Start.Station.Id) & !is.na(End.Station.Id)) |>
  group_by(Start.Station.Id, End.Station.Id) |>
  summarise(weight = n(), .groups = "drop") |>
  # Filter out trips with same start and end station
  filter(Start.Station.Id != End.Station.Id) |>
  # Sort by highest trip volume to lowest
  arrange(desc(weight)) |>
  # Keep top 100 rows
  slice_head(n = 100)

# Convert edge list to an igraph Network Object
bike_network <- graph_from_data_frame(edges, directed = TRUE)

# Run Louvain clustering
undirected_net <- as.undirected(bike_network, mode = "collapse", edge.attr.comb = "sum")
communities <- cluster_louvain(undirected_net)

# plot network
plot(communities, undirected_net, 
     vertex.size = 8,                
     vertex.label.cex = 0.7,         
     vertex.label.color = "black",
     edge.arrow.size = 0.3,
     # Scale the thickness of the connection lines based on the trip count
     edge.width = E(undirected_net)$weight / max(E(undirected_net)$weight) * 5,
     main = "Top 100 Routes: Bike Station Communities")


```

```{r}
# Extract which community each station belongs to
station_communities <- data.frame(
  station_id = communities$names,
  Community_ID = as.factor(communities$membership)
)

# Join the Community IDs with the spatial data (Lat/Long)
mapped_communities <- station_communities |>
  inner_join(stations_data, by = "station_id") |>
  select(station_id, name, lat, lon, Community_ID) |>
  filter(!is.na(lat) & !is.na(lon))

# create colour palette for different communities
pal_communities <- colorFactor(palette = "Set1", domain = mapped_communities$Community_ID)


# Generate map with leaflet
leaflet(mapped_communities) |>
  addTiles() |>
  setView(lng = -79.3832, lat = 43.6532, zoom = 12) |>
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    radius = 8, 
    fillColor = ~pal_communities(Community_ID),
    fillOpacity = 0.9,
    stroke = TRUE, 
    weight = 1,
    color = "black", 
    popup = ~paste("<b>Station:</b>", name, "<br>",
                   "<b>Network Community ID:</b>", Community_ID)
  ) |>
  addLegend("bottomright", pal = pal_communities, values = ~Community_ID,
            title = "Network Communities<br>(Based on Top 100 Routes)",
            opacity = 1)


```

```{r}
#  Prepare oubound Data
outflow_graph <- df |>
  filter((Start.Station.Id %in% beaches_stations) & !(End.Station.Id %in% beaches_stations)) |>
  count(Start.Station.Name, End.Station.Name, sort = TRUE) |>
  rename(from = Start.Station.Name, to = End.Station.Name, weight = n) |>
  
  # take only top 20 for visual simplicity
  slice_head(n = 20) |> 
  as_tbl_graph()

# Prepare inbound data
inflow_graph <- df |>
  filter(!(Start.Station.Id %in% beaches_stations) & (End.Station.Id %in% beaches_stations)) |>
  count(Start.Station.Name, End.Station.Name, sort = TRUE) |>
  rename(from = Start.Station.Name, to = End.Station.Name, weight = n) |>
  
  # again keeping top 20
  slice_head(n = 20) |> 
  as_tbl_graph()

# plotting function
plot_flow <- function(graph_data, title, color_edge) {
  ggraph(graph_data, layout = "fr") + 
    geom_edge_fan(aes(alpha = weight), color = color_edge, 
                  arrow = arrow(length = unit(3, 'mm')), 
                  end_cap = circle(3, 'mm')) + 
    
    geom_node_point(size = 4, color = "black") +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    
    theme_void() + 
    labs(title = title, subtitle = "Top 20 Connections by Volume")
}

# Generate outflow and inflow plots
p1 <- plot_flow(outflow_graph, "Outflow: Where do they go?", "firebrick")

p2 <- plot_flow(inflow_graph, "Inflow: Where do they come from?", "steelblue")

print(p1)
print(p2)


```

## Misc: User Type Check

Look at difference in distribution between annual vs casual members.

```{r}
# first remove outliers via 1.5*IQR method
# Calculate the quartiles
Q1 <- quantile(df$Trip..Duration, 0.25, na.rm = TRUE)
Q3 <- quantile(df$Trip..Duration, 0.75, na.rm = TRUE)
IQR_value <- Q3 - Q1

# Define the bounds 
lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

# Create a new filtered dataset
beaches_trips_no_outliers <- df |>
  filter(Trip..Duration >= lower_bound & Trip..Duration <= upper_bound) 

# Print how many rows were removed
rows_removed <- nrow(df) - nrow(beaches_trips_no_outliers)
print(paste("Removed", rows_removed, "outlier trips."))

# visualize distributions
ggplot(beaches_trips_no_outliers, aes(x = Trip..Duration, fill = User.Type)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(labels = scales::comma) + 
  labs(
    title = "Validation: Duration vs. User Type (Outliers Removed)",

    x = "Trip Duration (Seconds)",
    fill = "User Type",
    y = "Density"
  ) +
  theme_minimal()
```
