---
title: Modelling Bike Activity within The Beach BIA
date: today
date-format: long
format: pdf
execute:
  echo: false
---


```{r}
model_data_2025 <- readRDS(here("07_ExplanatoryModelling", "model_data_2025.rds"))
```

# Generalized Additive Models (Frequentist)

## The Baseline GLM

We fit a baseline Negative Binomial Generalized Linear Model (GLM) using only weather and calendar variables, intentionally ignoring the passing of time.

The Observation: The resulting ACF plot shows severe temporal autocorrelation. Lags 1 through 7 (an entire week) consistently spike well above the blue dashed significance thresholds.

The Implication: This proves that yesterday's unmeasured traffic heavily influences today's traffic. Our daily observations are mathematically not independent.

The Conclusion: We have definitive, visual justification for upgrading to a time-aware model. If we presented this GLM to the BIA, our standard errors would be artificially shrunken and our confidence intervals would be dangerously overconfident.


```{r}
library(mgcv)

# 1. Fit the strict parametric GLM (straight lines only, no time awareness)
fit_glm_nb <- gam(
  total_trips ~ temp_max + 
                is_precip + 
                precip_hour_count + 
                day_type_holiday_sum,
  family = nb(), 
  data = model_data_2025
)

# 2. View the basic coefficients
summary(fit_glm_nb)

# 3. Save the AIC score (we want this number to go down in our next steps)
cat("Baseline GLM AIC:", AIC(fit_glm_nb), "\n")
```

```{r}
# Plot the residuals of the baseline GLM
acf(residuals(fit_glm_nb, type = "pearson"), main = "ACF: Baseline GLM (No Time Correction)")
```

```{r}
library(car)
# We run VIF on your baseline GLM from earlier
car::vif(fit_glm_nb)
```

```{r}
library(tidyverse)
cor(model_data_2025 %>% dplyr::select(temp_max, usable_light_hours, is_precip, total_precip_mm, precip_hour_count))
```


## Time Series GAM using s(time_index)

```{r}
library(mgcv)

# 1. Strip the ordered factor
model_data_2025 <- model_data_2025 %>%
  mutate(dow = factor(dow, ordered = FALSE))

# 2. Fit the Time-Sponge GAM
fit_gam_time_sponge <- gam(
  total_trips ~ s(temp_max) + 
                usable_light_hours + 
                is_precip + 
                precip_hour_count + 
                day_type + 
                s(time_index, k = 50), # The Time Sponge!
  family = nb(), 
  data = model_data_2025,
  method = "REML"
)

# 3. Check the beautiful, non-polynomial coefficients
summary(fit_gam_time_sponge)

# 4. Verify it killed the autocorrelation
acf(residuals(fit_gam_time_sponge, type = "pearson"), main = "ACF: GAM with Time Sponge")
```



## GAM with explicti AR(1)

We avoid using the Generalized Additiive Mixed Model approach (GAMM) because it gives us untrustworthy p-values, due to the approximation method used for Negative Binomial.

However, using Day of the Week, improves its ACF by a lot.

```{r}
#| eval: false
library(mgcv)

model_data_2025 <- model_data_2025 %>%
  mutate(dow = factor(dow, ordered = FALSE))

# Fit the GAMM with an explicit AR(1) correlation structure
fit_gamm_ar1 <- gamm(
  total_trips ~ s(temp_max) + 
                usable_light_hours + 
                is_precip + 
                precip_hour_count + 
                dow,
  family = nb(), 
  data = model_data_2025,
  correlation = corAR1(form = ~ time_index) # Explicitly handling the temporal overlap
)

# 1. View the summary of the fixed effects
summary(fit_gamm_ar1$gam)

acf(residuals(fit_gamm_ar1$lme, type = "normalized"), main = "ACF: GAMM with corAR1")
```

## Lagged Response Variable GAM

By using Day of the Week, it improves its ACF by a lot, as it gets rid of the spikes at lags of 7.

```{r}
library(dplyr)
library(mgcv)

# 1. Engineer the Lagged Response Variable
model_data_2025 <- model_data_2025 %>%
  arrange(date) %>%
  mutate(
    # How many people biked yesterday?
    lag_total_trips = lag(total_trips, n = 1),
    dow_mon = fct_relevel(dow, "Mon")
  )

# 2. Fit the Dynamic GAM 
fit_gam_dynamic <- gam(
  total_trips ~ s(temp_max) + 
                usable_light_hours + 
                precip_hour_count + 
                dow_mon + 
                lag_total_trips, # This completely replaces corAR1
  family = nb(), 
  data = model_data_2025,
  method = "REML"                
)

# 3. View the business coefficients
summary(fit_gam_dynamic)

# 4. Check the ACF 
acf(residuals(fit_gam_dynamic, type = "pearson"), na.action = na.pass, main = "ACF: Dynamic GAM (Lagged Response)")
```


## Deciding on Rain predictor

```{r}
# Fit Model A: Volume
fit_mm <- gam(total_trips ~ dow_mon + is_summer_holiday + 
              s(temp_max, by = is_leisure_day) + s(time_index) + 
              total_precip_mm, 
              data = model_data_2025, family = nb(link = "log"))

# Fit Model B: Duration
fit_hours <- gam(total_trips ~ dow_mon + is_summer_holiday + 
                 s(temp_max, by = is_leisure_day) + s(time_index) + 
                 precip_hour_count, 
                 data = model_data_2025, family = nb(link = "log"))


# Compare them!
AIC(fit_mm, fit_hours)
```







```{r}
library(mgcv)

# ==========================================
# 1. FIT THE POISSON GAM
# ==========================================
fit_gam_poisson <- gam(
  total_trips ~ 
    dow_mon + 
    is_summer_holiday + 
    precip_hour_count +
    s(temp_max, by = is_leisure_day) + 
    s(time_index),
    
  data = model_data_2025,
  family = poisson(link = "log"),
  method = "REML"
)

# ==========================================
# 2. FIT THE NEGATIVE BINOMIAL GAM
# ==========================================
fit_gam_nb <- gam(
  total_trips ~ 
    dow_mon + 
    is_summer_holiday + 
    precip_hour_count +
    s(temp_max, by = is_leisure_day) + 
    s(time_index),
    
  data = model_data_2025,
  family = nb(link = "log"), # This is the only difference!
  method = "REML"
)

# ==========================================
# 3. THE OVERDISPERSION TEST
# ==========================================
# Calculate the sum of squared Pearson residuals from the Poisson model
pearson_sq <- sum(residuals(fit_gam_poisson, type = "pearson")^2)

# Divide by the residual degrees of freedom
dispersion_ratio <- pearson_sq / fit_gam_poisson$df.residual

cat("\n--- OVERDISPERSION TEST ---\n")
cat("Dispersion Ratio:", round(dispersion_ratio, 3), "\n")
cat("(Remember: > 1.5 means you have severe overdispersion and should switch)\n\n")

# ==========================================
# 4. THE AIC DUEL
# ==========================================
cat("--- AIC COMPARISON (Lower is Better) ---\n")
AIC(fit_gam_poisson, fit_gam_nb)
```






