---
title: Modelling Bike Activity within The Beach BIA
date: today
date-format: long
format: pdf
execute:
  echo: false
---


```{r setup}

library(here)
library(dplyr)
library(forcats)
library(brms)
library(mgcv)
library(bayesplot)
library(ggplot2)
library(posterior)
select <- dplyr::select
```


```{r}
model_data_2025 <- readRDS(here("07_ExplanatoryModelling", "model_data_2025.rds"))
```

```{r}
cor(model_data_2025 %>% dplyr::select(temp_max, usable_light_hours, is_precip, total_precip_mm, precip_hour_count))
```

```{r}
fit_bayes_nb <- readRDS("beach_bayes_nb_model.rds")
fit_bayes_poisson <- readRDS("beach_bayes_poisson_model.rds")
```


# Bayesian Model - Bayesian GAMM (Generalized Additive Mixed Model)

The correlation for maximum temperature and usable light hours in a day seemed highly correlated (0.76), so we opted to stick to temperature, to capture the summer rush of activity.


### Poisson Bayesian Model

```{r}
#| eval: false

# The Final Bayesian Model
fit_bayes_poisson <- brm(
  total_trips ~ 
    # Interaction: Fits two separate temp curves (one for workdays, one for leisure)
    s(temp_max, by = is_leisure_day) + # Solves the missing peaks in PPC
    
    # Macro-Trend: Captures the slow rise and fall of the tourist season
    s(time_index) +                    
    
    # Weekly Cycle: Gives every day its own baseline
    dow_mon +                          # Solves the weekly ACF spikes
    
    # Event Spikes & Rain
    is_summer_holiday + 
    precip_hour_count + 
    
    # Accounts for the daily ACF spikes
    ar(time = time_index, p = 1),      
    
  data = model_data_2025,
  family = poisson(), 
  cores = 4,      
  iter = 4000,    
  warmup = 1000,  
  control = list(adapt_delta = 0.99, max_treedepth = 15) 
)

# Rerun your Bayesian diagnostics to see the improvement
pp_check(fit_bayes_poisson, ndraws = 50)
acf(residuals(fit_bayes_poisson)[, "Estimate"], main = "New Bayesian ACF")
```


### Negative Binomial Bayesian Model

```{r}
#| eval: false
fit_bayes_nb <- brm(
  total_trips ~ 
    s(temp_max, by = is_leisure_day) + 
    s(time_index) + 
    dow_mon + 
    is_summer_holiday + 
    precip_hour_count + 
    ar(time = time_index, p = 1), 
    
  data = model_data_2025,
  family = negbinomial(), # The crucial fix
  cores = 4,      
  iter = 4000,    
  warmup = 1000,  
  control = list(adapt_delta = 0.99, max_treedepth = 15) 
)

summary(fit_bayes_nb)
```


### Comparison Check 

```{r}
# Add LOO criteria to both models
fit_bayes_nb <- add_criterion(fit_bayes_nb, "loo")
fit_bayes_poisson <- add_criterion(fit_bayes_poisson, "loo")

# Let them fight (A negative difference means the top model is better)
loo_compare(fit_bayes_nb, fit_bayes_poisson)
```


## MODEL CHECKING


```{r}
fit_bayes_final <- fit_bayes_nb
```


```{r}
# ==========================================
# 1. MCMC HEALTH & CONVERGENCE ASSUMPTIONS
# ==========================================
# Assumption: The chains have mixed properly and explored the posterior space.

# Overall summary (Check that all Rhat values are 1.00 and ESS > 1000)
summary(fit_bayes_final)

# Visual Trace Plots (Looking for "fuzzy caterpillars" with no wandering)
mcmc_plot(fit_bayes_final, type = "trace")

# Rhat Check (Should be a tight cluster exactly at 1.0)
mcmc_plot(fit_bayes_final, type = "rhat")
```



```{r}
color_scheme_set("viridis")
# 1. Create a "dictionary" linking the raw names to your clean report names
clean_titles <- c(
  "b_is_summer_holidayTRUE" = "Impact of Summer Holidays",
  "b_precip_hour_count"     = "Impact of Precipitation (Hours)"
)

# 2. Generate the plot and apply the clean names
mcmcplot <- mcmc_plot(fit_bayes_final, 
          type = "trace", 
          variable = names(clean_titles)) +
  
  # Overwrite the facet panels with your clean titles
  # (scales = "free_y" ensures the y-axes don't get squished together)
  facet_wrap(~parameter, scales = "free_y", labeller = as_labeller(clean_titles)) +
  
  # Clean up the main titles and theme
  labs(title = "MCMC Trace Plots for Select Predictors",
       subtitle = "Confirming chain convergence") +
  theme_minimal() + # Drops the heavy grey background
  theme(strip.text = element_text(size = 12, face = "bold"))

mcmcplot
```


```{r}
# ==========================================
# 2. POSTERIOR PREDICTIVE CHECKS (FIT ASSUMPTIONS)
# ==========================================
# Assumption: A Poisson distribution actually fits the data (no severe overdispersion).

color_scheme_set("blue")

# A. Density Overlay: Overall Fit
pp_check(fit_bayes_final, ndraws = 100) + 
  labs(
    title = "Posterior Predictive Check: Overall Distribution",
    subtitle = "Comparing actual daily trips (dark line) to 100 model simulations (light blue lines)",
    x = "Total Trips",
    y = "Density"
  ) +
  theme_minimal()

# B. Maximum Value Check: Extreme Highs
# Added binwidth = 50 to clear the stat_bin() warning. 
# (Adjust to 100 if the bars still look too thin!)
pp_check(fit_bayes_final, type = "stat", stat = "max", ndraws = 100, bins = 15) + 
  labs(
    title = "Model Validation: Extreme Peak Days",
    subtitle = "Does the model's predicted maximum match the actual observed maximum?",
    x = "Maximum Trips in a Single Day",
    y = "Frequency in Simulations"
  ) +
  theme_minimal()

# C. Zero or Low Value Check: Extreme Lows
# Added binwidth = 1 because minimum trips are usually small discrete integers (0, 1, 2, etc.)
pp_check(fit_bayes_final, type = "stat", stat = "min", ndraws = 100, binwidth = 1) +
  labs(
    title = "Model Validation: Low Activity Days",
    subtitle = "Checking if the model accurately captures the quietest days of the year",
    x = "Minimum Trips in a Single Day",
    y = "Frequency in Simulations"
  ) +
  theme_minimal()
```



```{r}
# ==========================================
# 3. RESIDUALS & AUTOCORRELATION (INDEPENDENCE ASSUMPTION)
# ==========================================
# Assumption: The AR(1) and dow_mon terms removed the temporal patterns.

# Extract Bayesian residuals (median estimates)
bayes_resids <- residuals(fit_bayes_final)[, "Estimate"]

# Plot ACF to ensure Lags 1, 7, 14, 21 are within the blue dashed lines
acf(bayes_resids, main = "ACF of Bayesian Residuals", lag.max = 28)

```


```{r}

# view proportion of sampled coefficients that are positive or negative
posterior_samples <- as_draws_df(fit_bayes_final)

# view what proportion of rain covariates are negative
# if 100% then know rain always results in less riders
head(select(posterior_samples, b_Intercept, b_precip_hour_count))

prob_rain_negative <- sum(posterior_samples$b_precip_hour_count < 0) / nrow(posterior_samples)

cat("There is a", prob_rain_negative * 100, "% probability that rain reduces traffic.\n")


prob_sun_busier <- sum(posterior_samples$b_dow_monSun > posterior_samples$b_dow_monSat) / nrow(posterior_samples)

cat("There is a", prob_sun_busier * 100, "% probability that Sundays are busier than Saturdays.\n")
```





## Visualizations of Results



```{r}
# This extracts the marginal effects for your splines and linear terms
bayesian_plots <- conditional_effects(fit_bayes_final)

# Because it uses ggplot2 under the hood, you can plot it instantly
plot(bayesian_plots, ask = FALSE)
```


```{r}
library(ggplot2)
library(viridis)

# Extract data from the conditional_effects object
# (Assuming the order: 1 = precip, 2 = day_category, 3 = temp_max)
data_precip <- as.data.frame(bayesian_plots$precip_hour_count)
data_day    <- as.data.frame(bayesian_plots$dow_mon)
data_temp   <- as.data.frame(bayesian_plots$temp_max)
data_interaction <- as.data.frame(bayesian_plots[["temp_max:is_leisure_day"]])
data_time <- as.data.frame(bayesian_plots$time_index)

# Common theme setup
beach_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# Plot 1: Precipitation
p1 <- ggplot(data_precip, aes(x = precip_hour_count, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), fill = "steelblue", alpha = 0.2) +
  geom_line(color = "steelblue", size = 1.2) +
  labs(title = "Impact of Precipitation on Biking Volume",
       x = "Hours of Precipitation", y = "Estimated Daily Trips") +
  beach_theme

# Plot 2: Day Category
p2 <- ggplot(data_day, aes(x = dow_mon, y = estimate__)) +
  geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.2, color = "grey30") +
  geom_point(size = 4, color = "darkred") +
  labs(title = "Ridership by Day Category",
       x = "Day Type", y = "Estimated Daily Trips") +
  beach_theme

# Plot 3: Temperature
p3 <- ggplot(data_temp, aes(x = temp_max, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), fill = "orange", alpha = 0.2) +
  geom_line(color = "orange3", size = 1.2) +
  labs(title = "Temperature-Driven Activation",
       x = "Maximum Temperature (°C)", y = "Estimated Daily Trips") +
  beach_theme

p_interaction <- ggplot(data_interaction, aes(x = temp_max, y = estimate__, color = is_leisure_day, fill = is_leisure_day)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.2, color = NA) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("Leisure" = "darkorange", "Workday" = "steelblue")) +
  scale_fill_manual(values = c("Leisure" = "darkorange", "Workday" = "steelblue")) +
  labs(title = "Weekend + Holiday Multiplier: Temperature Effect by Day Type",
       x = "Maximum Temperature (°C)", 
       y = "Estimated Daily Trips",
       color = "Day Type", fill = "Day Type") +
  beach_theme +
  theme(legend.position = "bottom")


p_time <- ggplot(data_time, aes(x = time_index, y = estimate__)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), fill = "seagreen", alpha = 0.2) +
  geom_line(color = "seagreen4", size = 1.2) +
  labs(title = "Underlying Seasonal Trend (Independent of Weather)",
       x = "Day of the Year (Time Index)", 
       y = "Baseline Daily Trips") +
  beach_theme


# Display them
p1; p2; p3; p_interaction; p_time
```



### Coefficients

```{r}
# Extract the summary table
mcmc_summary <- fixef(fit_bayes_final)

# Convert the log estimates to readable multipliers (IRRs)
mcmc_multipliers <- exp(mcmc_summary)

# View the clean multipliers for your report
round(mcmc_multipliers, 2)
```


```{r}
library(dplyr)
library(gt)
library(tibble)

# 1. Extract and Exponentiate directly from the model
# (Using fit_bayes_nb as it was the winner in your LOO test)
mcmc_summary <- fixef(fit_bayes_nb) %>% 
  as.data.frame() %>% 
  exp() %>%                # Convert logs to multipliers (IRRs)
  rownames_to_column("term_raw")

# 2. Define a mapping for your labels
# This tells R exactly which raw variable name corresponds to your report name
label_map <- c(
  "dow_monSat" = "Saturday",
  "dow_monSun" = "Sunday",
  "is_summer_holidayTRUE" = "Summer Holiday",
  "precip_hour_count" = "Precipitation (per hour)"
)

# 3. Filter and Transform the data
formatted_table_data <- mcmc_summary %>%
  filter(term_raw %in% names(label_map)) %>%
  mutate(
    Predictor = label_map[term_raw],
    # Calculate percentages
    pct_change = (Estimate - 1) * 100,
    lower_ci = (Q2.5 - 1) * 100,
    upper_ci = (Q97.5 - 1) * 100,
    
    # Categorize direction
    Direction = case_when(
      pct_change > 0 ~ "Increase",
      pct_change < 0 ~ "Decrease",
      TRUE ~ "No Change"
    ),
    
    # Format the CI string
    `95% Credible Interval` = sprintf("[ %+.1f%% , %+.1f%% ]", lower_ci, upper_ci)
  ) %>%
  select(Predictor, Direction, `Estimated Change (%)` = pct_change, `95% Credible Interval`)

# 4. Generate the gt table
formatted_table_data %>%
  gt() %>%
  tab_header(
    title = md("**Expected Change in Casual Ridership**"),
    subtitle = "Relative to a baseline standard Monday"
  ) %>%
  fmt_number(
    columns = `Estimated Change (%)`,
    decimals = 1,
    pattern = "{x}%"
  ) %>%
  cols_align(
    align = "center",
    columns = c(Direction, `Estimated Change (%)`, `95% Credible Interval`)
  ) 
  #%>%
  #tab_source_note(
   # source_note = "Note: Values extracted directly from Bayesian Negative Binomial posterior #distributions." )
```



```{r}
# Calculate the probability of direction for your core variables
prob_sat_positive <- sum(posterior_samples$b_dow_monSat > 0) / nrow(posterior_samples)
prob_sun_positive <- sum(posterior_samples$b_dow_monSun > 0) / nrow(posterior_samples)
prob_holiday_positive <- sum(posterior_samples$b_is_summer_holidayTRUE > 0) / nrow(posterior_samples)
prob_rain_negative <- sum(posterior_samples$b_precip_hour_count < 0) / nrow(posterior_samples)

# Print them out nicely
cat(
  "Probability Saturday is busier than Monday:     ", prob_sat_positive * 100, "%\n",
  "Probability Sunday is busier than Monday:       ", prob_sun_positive * 100, "%\n",
  "Probability Holidays are busier than Mondays:   ", prob_holiday_positive * 100, "%\n",
  "Probability Rain decreases traffic:             ", prob_rain_negative * 100, "%\n"
)
```

```{r}

# Generate the effects object
eff <- conditional_effects(fit_bayes_nb, "temp_max:is_leisure_day")

# Extract the underlying dataframe
plot_data <- as.data.frame(eff[[1]])

# Let's find the values specifically at 30°C to label them
labels_30deg <- plot_data %>%
  # Find the row closest to 30 degrees for both groups
  group_by(is_leisure_day) %>%
  filter(abs(temp_max - 30) == min(abs(temp_max - 30)))

ggplot(plot_data, aes(x = temp_max, y = estimate__, color = is_leisure_day)) +
  # The Main Curves
  geom_line(size = 1.2) +
  # Adding the points and labels at 30°C
  geom_point(data = labels_30deg, size = 4) +
  geom_label(data = labels_30deg, 
             aes(label = paste0(round(estimate__), " trips")),
             nudge_y = 40,  # Move label up slightly
             fontface = "bold",
             show.legend = FALSE) +
  # Aesthetics
  scale_color_manual(values = c("Leisure" = "darkorange", "Workday" = "steelblue")) +
  labs(title = "The 30°C Activation Gap",
       subtitle = "Leisure days see a massive surge compared to workdays at peak heat",
       x = "Maximum Temperature (°C)", 
       y = "Expected Daily Trips",
       color = "Day Type") +
  beach_theme
```





## Interpretations and Insights


Slide Title: Drivers of Bike Share Traffic at the Beach BIA
Baseline Reference: A standard weekday (Monday–Friday) with zero precipitation.

📅 The Calendar Surges (Human Behavior)
Compared to a standard weekday, weekend and holiday leisure traffic dramatically reshapes trail usage. We have roughly 95% statistical certainty of the following ranges:

Saturdays: Drive an estimated +33.6% surge in ridership (Expected range: +17% to +52%).

Sundays: Drive an estimated +53.7% surge in ridership (Expected range: +35% to +75%).

Statutory Holidays: Are the absolute peak, driving a massive +120.3% surge in ridership, literally more than doubling standard weekday traffic (Expected range: +69% to +182%).

🌧️ The Weather Penalty (Precipitation)
Rain duration is a highly credible deterrent to local traffic.

The Hourly Cost: Every single hour of recorded precipitation decreases that day's total ridership by an estimated 6.8%.

Business Certainty: We are virtually 100% certain that rain has a negative impact, with the true penalty falling strictly between -5.9% and -8.6% per hour.

☀️ The Temperature Curve
(Placeholder for the visual plot)

Temperature does not operate on a straight line. The model confirms a highly credible non-linear "hinge" effect, where ridership accelerates rapidly as weather warms, before plateauing at peak summer temperatures.



# User Type Segmentation

```{r}
model_data_long <- model_data_2025 %>%
  pivot_longer(
    cols = c(trips_casual, trips_member), # Or whatever your column names are
    names_to = "user_type",
    values_to = "trips"
  ) %>%
  mutate(user_type = factor(user_type, 
                          levels = c("trips_casual", "trips_member")))#,
                          #labels = c("Casual", "Annual")))
```


```{r}
# The "Champion" Interaction Model
fit_user_split <- brm(
  trips ~ s(temp_max, by = user_type) +    # Unique curves for each group
          user_type * precip_hour_count +  # Rain sensitivity by group
          user_type * day_category +       # Weekend/Holiday behavior by group
          ar(time = time_index, gr = user_type, p = 1), 
  data = model_data_long,
  family = poisson(),
  cores = 4, 
  iter = 4000,
  control = list(adapt_delta = 0.99)
)
```


```{r}

saveRDS(fit_user_split, "fit_user_split.rds")
summary(fit_user_split)
```

```{r}
library(ggplot2)

# Extract and plot the temperature effect
eff_temp <- conditional_effects(fit_user_split, effects = "temp_max:user_type")

plot(eff_temp, 
     points = FALSE, 
     theme = theme_minimal())[[1]] + 
  labs(title = "Temperature Sensitivity: Annual vs. Casual",
       subtitle = "Leisure ridership accelerates faster in high heat",
       x = "Max Temperature (°C)", 
       y = "Predicted Daily Trips")
```

Temperature Impact by User Type
Cold Weather Resilience: Annual members maintain a consistent ridership baseline in sub-zero temperatures while casual activity remains near zero.

Thermal Acceleration: Both groups exhibit exponential growth as weather warms, with members driving higher absolute volumes during peak summer heat.




```{r}
# Extract the interaction between rain and user type
eff_rain <- conditional_effects(fit_user_split, effects = "precip_hour_count:user_type")

plot(eff_rain, theme = theme_minimal())[[1]] +
  labs(title = "Rain Resilience: Annual vs. Casual",
       subtitle = "Annual members (commuters) are significantly more weather-proof",
       x = "Hours of Precipitation",
       y = "Predicted Daily Trips") +
  scale_color_manual(values = c("annual" = "#2c3e50", "casual" = "#e74c3c")) +
  scale_fill_manual(values = c("annual" = "#2c3e50", "casual" = "#e74c3c"))
```

Leisure vs. Utility: Casual riders cancel optional trips instantly while annual members maintain essential daily commutes.

Volume Stability: Casual traffic faces a near-total collapse during rain, whereas members retain a 30% ridership floor.

Operational Backbone: Annual members provide a weather-proof foundation for Beach BIA docks regardless of conditions.


```{r}
# Extract the interaction between the calendar and user type
eff_calendar <- conditional_effects(fit_user_split, effects = "day_category:user_type")

plot(eff_calendar, theme = theme_minimal())[[1]] +
  labs(title = "Calendar Behavior by User Type",
       subtitle = "Weekends and Holidays drive Casual growth, not Member growth",
       x = "Day Category",
       y = "Predicted Daily Trips") +
  scale_color_manual(values = c("annual" = "#2c3e50", "casual" = "#e74c3c"))
```



Weekday Foundation: Annual members provide the system's baseline, taking over double the trips of casual riders during standard work weeks.

Leisure Scaling: Casual ridership triples between weekdays and holidays, rapidly closing the volume gap during peak leisure periods.

Holiday Parity: Statutory holidays represent peak system demand where casual and annual ridership reach nearly identical high-volume levels.




```{r}
library(ggplot2)
library(dplyr)

# 1. Define the Beach BIA Palette
# Annual = Deep Water Blue, Casual = Sand/Sunset Gold
beach_cols <- c("Annual" = "#1A5276", "Casual" = "#D4AC0D")

# 2. Extract Data
data_temp_int <- as.data.frame(eff_temp[[1]])
data_rain_int <- as.data.frame(eff_rain[[1]])
data_calendar_int <- as.data.frame(eff_calendar[[1]])

# 3. Custom Theme
beach_theme_fancy <- theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, family = "sans"),
    plot.subtitle = element_text(size = 12, color = "grey40"),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )

# Plot A: Temperature Sensitivity
p_temp <- ggplot(data_temp_int, aes(x = temp_max, y = estimate__, color = user_type, fill = user_type)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.15, color = NA) +
  geom_line(size = 1.3) +
  scale_color_manual(values = beach_cols) +
  scale_fill_manual(values = beach_cols) +
  labs(title = "Temperature Sensitivity by User Group",
       #subtitle = "Casual ridership shows a delayed increased in trips compared to stable annual growth",
       x = "Maximum Daily Temperature (°C)", y = "Predicted Daily Trips") +
  beach_theme_fancy

# Plot B: Rain Resilience
p_rain <- ggplot(data_rain_int, aes(x = precip_hour_count, y = estimate__, color = user_type, fill = user_type)) +
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = 0.15, color = NA) +
  geom_line(size = 1.3) +
  scale_color_manual(values = beach_cols) +
  scale_fill_manual(values = beach_cols) +
  labs(title = "Impact of Precipitation by User Type",
       #subtitle = "Annual members maintain higher activity levels compared to casual",
       x = "Hours of Precipitation", y = "Predicted Daily Trips") +
  beach_theme_fancy


# 3. Beautified Interaction Plot
p_calendar <- ggplot(data_calendar_int, aes(x = day_category, y = estimate__, color = user_type)) +
  # Use error bars for the 95% Credible Intervals
  geom_errorbar(aes(ymin = lower__, ymax = upper__), width = 0.2, size = 1, alpha = 0.7, position = position_dodge(0.4)) +
  # Large points for the estimates
  geom_point(size = 4, position = position_dodge(0.4)) +
  scale_color_manual(values = beach_cols) +
  labs(title = "Day of Week and User Type Relationship on Bike Trips",
       #subtitle = "Weekends and Holidays enables Casual riders, while Annual volume remains stable",
       x = "Day Category", 
       y = "Predicted Daily Trips") +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11, color = "grey40"),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.major.x = element_blank() # Clean up the vertical lines
  )

# Display
p_calendar
# Display
p_temp
p_rain
```









